---
title: "CapstoneProject"
author: "Hans M. Rupp"
date: "26 5 2022"
output: pdf_document
---

```{r}
library(tidyverse)
library(caret)
library(gridExtra)
```


TODO: Replace later with code from project site
```{r}
load(file="data/edx")
load(file="data/validation")
load(file="data/movies")
```
Overview
```{r}
str(edx)

```
```{r echo=FALSE}
edx_rows <- nrow(edx)
edx_cols <- ncol(edx)
val_rows <- nrow(validation)
val_cols <- ncol(validation)
```

The edx dataset contains `r edx_rows` observations (rows) with `r edx_cols` features (columns). It will be used for training the model.
The validation dataset contains `r val_rows` observations (rows) with `r val_cols` features (columns). It will be used for validating the model

Checking for missing values
```{r}
nas <- function(x) { any(is.na(x)) }
edx %>% summarise_all(nas)
validation %>% summarise_all(nas)
```
There are no columns with NAs in the edx or the validation data. So no clean-up is necessary

Overall distribution of the ratings

```{r}
edx %>% ggplot(aes(x=rating)) + geom_histogram(binwidth = 0.5) + scale_x_continuous(breaks = seq(min(edx$rating), max(edx$rating), by = 0.5))

```

The different genres are
```{r}
dist_genres <- edx %>% distinct(genres)
dg <- as.vector(str_split(dist_genres$genres, "\\|", simplify = T))
unique(dg[dg != ""])
```


Distribution of ratings for  different genres

```{r}
hist_drama <- edx %>% filter(str_detect(genres, "Drama")) %>% ggplot(aes(x=rating)) + geom_histogram(binwidth = 0.5) + scale_x_continuous(breaks = seq(0.5, 5, by = 1)) + labs(title="Drama")

hist_comedy <- edx %>% filter(str_detect(genres, "Comedy"))  %>% ggplot(aes(x=rating)) + geom_histogram(binwidth = 0.5) + scale_x_continuous(breaks = seq(0.5, 5, by = 1)) + labs(title="Comedy")

hist_thriller <- edx %>% filter(str_detect(genres, "Thriller"))  %>% ggplot(aes(x=rating)) + geom_histogram(binwidth = 0.5) + scale_x_continuous(breaks = seq(0.5, 5, by = 1)) + labs(title="Thriller")

hist_romance <- edx %>% filter(str_detect(genres, "Romance"))  %>% ggplot(aes(x=rating)) + geom_histogram(binwidth = 0.5) + scale_x_continuous(breaks = seq(0.5, 5, by = 1)) + labs(title="Romance")

grid.arrange(hist_drama, hist_comedy, hist_thriller,hist_romance, ncol = 2)
```

Different users have widely different numbers of ratings

```{r}
edx %>% count(userId) %>% ggplot(aes(x=n)) + geom_histogram() + scale_x_log10() + xlab("Ratings per user (log)")
```

Different users have different rating means

```{r}
edx %>% group_by(userId) %>% summarise(user_mean=mean(rating)) %>% ggplot(aes(x=user_mean)) + geom_histogram()  + xlab("Average ratings per user")
```


Some movies are rated more often

```{r}
edx %>% group_by(movieId) %>% summarise(n=n()) %>% ggplot(aes(x=n)) + geom_histogram() + scale_x_log10() + xlab("Ratings per movie (log)")

```

##Release year has some influence on average rating
Extract the release date from the title

```{r}
release_year_edx <- as.numeric(str_sub(edx$title, start=-5, end=-2))
edx <- edx %>% mutate(year=release_year_edx)
validation <- validation %>% mutate(year=as.numeric(str_sub(title, start=-5, end=-2)))
```
Plot of the average rating for each release year

```{r}
edx %>% group_by(year) %>% summarise(avg_rating = mean(rating)) %>% ggplot(aes(x=year, y=avg_rating)) + geom_point() + geom_smooth()

```



# Matrix Factorization
We will use Matrix Factorization to build a model to predict movie ratings

movie _i_
user _u_
year _y_


## RMSE
The Root Mean Square Error will be used as the Loss function

RMSE = $\sqrt{\frac{1}{N}\sum_{u,i,y} (\hat{y}_{u,i,y}-y_{u,i,y})^2}$
```{r}

RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
```

## Naive model: simply using the mean ratings
This model assumes the same rating $\mu$ for all movies i and users u with an error $\epsilon _{u,i}$  
Y~u,i~ = $\mu + \epsilon _{u,i}$

```{r}
mu_rating <- mean(edx$rating)
rmse_mean <- RMSE(validation$rating, mu_rating)
rmse_mean

```
This gives a RMSE of `r rmse_mean`

We build a table of the RMSEs for the different models:

```{r}
models <- tibble(model="Mean", rmse=rmse_mean)
```

## Effect of different movies
As we saw above different movies are rated differently.

We can add the term b~i~ for the average ranking for movie i

$Y_{u,i} = \mu + b_{i} + \epsilon_{u,i}$

Theoretically we could calculate b~i~ using linear regression. But with this amount of data that would be very costly.

Instead we can just use $b_{i} = Y_{u,i} - \mu$

Calculate the average rating b_i for each movie minus the overall average rating mu_rating = `r mu_rating`:

```{r}

movie_avgs <- edx %>% group_by(movieId) %>% summarise(b_i = mean(rating - mu_rating))
```
Plot them:

```{r}

movie_avgs %>% ggplot(aes(b_i)) + geom_histogram(bins = 10)
```

Join the movie_avg data.frame to the edx (training) and validation set:

```{r}

edx_movies <- edx %>% left_join(movie_avgs, by="movieId")
validation_movies <- validation %>% left_join(movie_avgs, by="movieId")
```
Calculating the predicted ratings $Y_{u,i} = \mu + b_{i}$

```{r}
predicted_ratings_movies <- mu_rating + validation_movies$b_i
```


Calculating the new RMSE

```{r}
rmse_movies <- RMSE(predicted_ratings_movies, validation$rating)
models <- rbind(models, c("movies", rmse_movies))
models
```

The RMSE has improved to `r models$rmse[2]` 

## Effect of different users
As we saw above different users have different rating means.

We can add the term b~u~ for the average rating for user u

$Y_{u,i} = \mu + b_{i} + b_{u} + \epsilon_{u,i}$

Calculate the average rating b_u for each movie minus the overall average rating mu_rating and b_i:

```{r}
user_avg <- edx_movies %>% group_by(userId) %>% summarise(b_u = mean(rating - mu_rating - b_i)) 
```
Join the user_avg data.frame to the edx_movies (training) and validation_movies set:

```{r}
edx_movies_users <- edx_movies %>% left_join(user_avg, by="userId")
validation_movies_users <- validation_movies %>% left_join(user_avg, by="userId")
```
Plot the distribution of user effects b_u

```{r}

edx_movies_users %>% ggplot(aes(x=b_u)) + geom_histogram()
```
Calculating the predicted ratings $Y_{u,i} = \mu + b_{i} + b_{u}$

```{r}

predicted_ratings_movies_users <- mu_rating + validation_movies_users$b_i + validation_movies_users$b_u
rmse_movies_users <- RMSE(predicted_ratings_movies_users, validation$rating)
models <- rbind(models, c("movies users", rmse_movies_users))
models
```
The RMSE has improved to `r models$rmse[3]` 

## Effect of the release year

We can add the term b~y~ for the average rating for movies released in year y

$Y_{u,i} = \mu + b_{i} + b_{u} + b_{y} + \epsilon_{u,i}$

Calculate the average rating b_u for each movie minus the overall average rating mu_rating and b_i and b_u:

```{r}
year_avg <- edx_movies_users %>% group_by(year) %>% summarise(b_y = mean(rating - mu_rating - b_i - b_u))

```
Plot the effect of the release year b_y

```{r}
year_avg %>% ggplot(aes(x=b_y)) + geom_histogram()
```
Join the year_avg data.frame to the edx_movies_users (training) and validation_movies_users set:

```{r}
edx_movies_users_years <- edx_movies_users %>% left_join(year_avg, by="year")
validation_movies_users_years <- validation_movies_users %>% left_join(year_avg, by="year")
```
Calculating the predicted ratings $Y_{u,i} = \mu + b_{i} + b_{u} + b_{y}$

```{r}
predicted_ratings_movies_users_years <- mu_rating + validation_movies_users_years$b_i + validation_movies_users_years$b_u + validation_movies_users_years$b_y

rmse_movies_users_years <- RMSE(predicted_ratings_movies_users_years, validation$rating)
rmse_movies_users_years

models <- rbind(models, c("movies users years", rmse_movies_users_years))
models
```

The RMSE has only marginally improved to `r models$rmse[4]`
